---
title: "Data Anaysis of Beetle Richness and Leaf Litter"
author: "Zachary krausman, Eryn Lackey, Riley Wares, and Afia Lockett"
date: "2026-03-24"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load in Packages

```{r load packages}

#install.packages("pacman")
library(pacman)

# Load packages (emmeans used for statistical tests)
pacman::p_load(tidyverse, here, cowplot, emmeans)

```

Import datasets from the accompanying "data_wrangle_official.Rmd" file

```{r import datasets}

litter_averages_total <- read.csv(here("data", "processed", "litter_averages_total.csv"))
beetle_abundance_taxonID <- read.csv(here("data", "processed", "beetle_abundance_taxonID.csv"))
beetle_richness_taxonID <- read.csv(here("data", "processed", "beetle_richness_taxonID.csv"))

```

Preliminary visuals of beetle and litter observations per month.

```{r check monthliness}

ggplot(data = litter_averages_total, 
       aes(x = monthCollected))+
  geom_dotplot(dotsize = 0.025)

ggplot(data = beetle_richness_taxonID, 
       aes(x = month(collectDate)))+
  geom_dotplot(dotsize = 0.025)

```

A note to future explorers: We do not have data for the winter (or
between march and october). We are choosing to focus on the rough
growing season (april to october)

For discussion: it would be interesting to see what is visible in the
winter, but requires more data.

Create average richness with the beetle data.

```{r create average richness}

beetle_richness_monthly <- beetle_richness_taxonID %>% 
  mutate(monthCollected = month(collectDate), 
         yearCollected = year(collectDate)) %>%
  group_by(plotID, yearCollected, monthCollected) %>%
  summarize(avg_richness_monthly = mean(richness))

```

Preparing beetle and litter data for joining including creating means.

```{r prelim join tables}

beetle_richness_by_site_month <- beetle_richness_monthly %>%
  mutate(siteID = substring(plotID, 0, 4)) %>%
  group_by(siteID, yearCollected, monthCollected) %>%
  summarize(avg_richness_plotmonth = mean(avg_richness_monthly))

litter_mass_by_site_month <- litter_averages_total %>%
  mutate(siteID = substring(plotID, 0, 4)) %>%
  group_by(siteID, yearCollected, monthCollected) %>%
  summarize(avg_mass_plotmonth = mean(avgmass), 
            avg_C_plotmonth = mean(avgCpercent),
            avg_N_plotmonth = mean(avgNpercent))
  
```

Joining litter and beetle data! Dropping Carbon and Nitrogen due to no
overlap with beetle data

```{r}

beetle_litter_joined_table <- beetle_richness_by_site_month %>%
  full_join(litter_mass_by_site_month) %>%
  drop_na("avg_richness_plotmonth","avg_mass_plotmonth") %>%
  select(siteID, yearCollected, monthCollected, avg_richness_plotmonth, avg_mass_plotmonth)

```

Linear Model for beetle richness as a function of average leaf litter
mass.

```{r}

beetle_litter_lm <- lm(avg_richness_plotmonth ~ (avg_mass_plotmonth), data= beetle_litter_joined_table) 

summary(beetle_litter_lm)

```

Checking assumptions for a linear model.

```{r}
par(mfrow = c(1, 3))

plot(beetle_litter_lm, which = 1:3)

shapiro.test(beetle_litter_lm$residuals)
```

Average leaf litter mass per month was not a significant predictor of
average beetle richness per month as seen by the linear regression
model's p-value much greater than 0.05 (F(1, 137)=.9278, p=.337). A
linear regression model was used as both average leaf litter mass per
month and average beetle richness per month are continuous variables.
However, the linear regression model did not met the required
assumptions of normality due to its low p-value of less than .05
produced by the Shapiro test and the strong concentration of points on
the right side of the residuals v fitted and scale location plots.
However, this lack of leaf litter mass as a predictor of beetle richness
leds us to wonder whether leaf litter may still be a predictor for
beetle abundance.

Average leaf litter mass per month was not a significant predictor of
average beetle richness per month as seen by the p-value much greater
than 0.05 **(F(1, 137) = 0.9278, p = 0.337)**.

------------------------------------------------------------------------

**DO NOT INCLUDE IN POSTER** Scatterplot of beetle richness and average
leaf litter mass.

```{r make a graph}

ggplot(beetle_litter_joined_table, 
       aes(x = avg_mass_plotmonth, 
           y = avg_richness_plotmonth,
           color = siteID)) +
  labs(x = "Average Leaf Litter Mass per Month (g)",
       y = "Average Beetle Richness per Month") + 
  geom_point(size = 1) +
  theme_bw()

```

Figure X. Average beetle richness per month as a function of the average
leaf litter mass per month (g). There was no significant relationship
between average beetle richness per month and average leaf litter mass
per month (g) **(F(1, 137) = 0.9278, p = 0.337)**. \*\*\* Adding natural
log fit on the beetle and litter data linear model.

```{r}

beetle_litter_lm_lin1 <- lm((avg_richness_plotmonth) ~ log(avg_mass_plotmonth), data= beetle_litter_joined_table) 

summary(beetle_litter_lm_lin1)

```

------------------------------------------------------------------------

**KEEPING THIS FOR THE POSTER** Checking assumptions for log fit linear
model.

```{r}
par(mfrow = c(1, 3))

plot(beetle_litter_lm_lin1, which = 1:3)

shapiro.test(beetle_litter_lm_lin1$residuals)
```

Scatterplot of log fit beetle richness and average leaf litter mass.

```{r make a graph}

ggplot(beetle_litter_joined_table, 
       aes(x = log(avg_mass_plotmonth), 
           y = avg_richness_plotmonth)) +
  labs(x = "Ln Average Leaf Litter Mass per Month (g)",
       y = "Average Beetle Richness per Month") + 
  geom_smooth(method = "lm", 
              col = "black") +
  geom_point(size = 1,
             aes(col = siteID)) +
  theme_bw()

```

**Figure 1.** Scatterplot showing the significant relationship between
the log of the average leaf litter mass per month and average beetle
richness per month. Trendline shows line of best fit globally. (F(1,
137) = 5.801, p = 0.01735). \*\*\* Exploratory anova analysis data based
on site.

```{r graph by site}

# aov() should be for factor/categorical predictors: testing whether means different among groups
beetle_litter_aov <- aov(avg_richness_plotmonth ~ avg_mass_plotmonth * siteID, data = beetle_litter_joined_table) 

summary(beetle_litter_aov)

plot(beetle_litter_aov, which = 1:3)
# Quantitative tests for normality, homogeneity among groups? 

```

------------------------------------------------------------------------

**DO NOT NEED FOR POSTER** Visual of anova model produced.

```{r}

# Not sure why this plot is using the fitted model as the data source?
ggplot(beetle_litter_aov, 
       aes(x = avg_mass_plotmonth, 
           y = avg_richness_plotmonth, 
           color = siteID)) +
  geom_point(size=2.0) +
  labs(x = "Average Leaf Litter Mass per Month (g)",
       y = "Average Beetle Richness per Month",
       col = "Site ID") +
  geom_smooth(method = "lm") +
  theme_bw()

## Remember you typically only include a trendline for statistically significant relationships

```

Figure XX. Scatterplot visualizing the relationship between average leaf
litter mass per month and average beetle richness per month based on
site.

Exploratory chunk using emmeans model plugin \*\*\* KF: Running
emmeans() command below gives output: "NOTE: Results may be misleading
due to involvement in interactions". As mentioned above, probably don't
want to include the interaction and stick to categorical variable in
your aov(). Can also use TukeyHSD() frmn base R if not trying to compare
adjusted means

```{r}

# This is another type of linear model using this plugin
em_beetle_model <- emmeans(beetle_litter_aov, ~ siteID)
tukey_beetle_litter <- contrast(em_beetle_model, method = "pairwise", adjust = "Tukey")
summary(tukey_beetle_litter, infer = TRUE, level = 0.95)

```

Adding aggregate abundance per month variable to joined table.

```{r}
beetle_abundance_agg <- beetle_abundance_taxonID %>%
  mutate(siteID = substring(plotID, 0, 4)) %>%
  group_by(siteID, collectDate) %>%
  summarize(agg_abundance = sum(count)) %>%
  mutate(monthCollected = month(collectDate), 
         yearCollected = year(collectDate)) %>%
  group_by(siteID, yearCollected, monthCollected) %>%
  summarize(avg_agg_abundance_monthly = mean(agg_abundance))
```

Joining beetle abundance per month with litter mass per month.

```{r}
beetle_abundance_litter_joined <- beetle_litter_joined_table %>%
  full_join(beetle_abundance_agg)
```

Linear model for abundance per month table, with log of litter mass
which makes it more normal, although still not normal.

```{r}
model_beetle_litter_abundance <- lm(avg_agg_abundance_monthly ~ log(avg_mass_plotmonth), data = beetle_abundance_litter_joined)

summary(model_beetle_litter_abundance)
```

------------------------------------------------------------------------

**INCLUDING THIS IN THE POSTER** Checking assumption for abundance
linear model.

```{r}
par(mfrow = c(1, 3))

plot(model_beetle_litter_abundance, which = 1:3)

shapiro.test(model_beetle_litter_abundance$residuals)

```

Scatterplot of log fit beetle richness and average leaf litter mass.

```{r make a graph}

ggplot(beetle_abundance_litter_joined, 
       aes(x = log(avg_mass_plotmonth), 
           y = avg_agg_abundance_monthly)) +
  labs(x = "Ln Average Leaf Litter Mass per Month (g)",
       y = "Aggergate Beetle Abundance per Month") + 
  geom_smooth(method = "lm", 
              col = "black") +
  geom_point(size = 1,
             aes(col = siteID)) +
  labs(col = "Site") +
  theme_bw()
```

**Figure 2.** Aggregate beetle abundance per month as a function of log
of the average leaf litter mass per month (g). Trendline shows global
best fit. **(F(1, 137) = 4.066, p = 0.04572)** There is a slight but
significant negative relationship between the log of the average leaf
litter mass per month and aggregate beetle abundance per month.

Export joined beetle richness and average leaf litter data.

```{r send it out}

beetle_litter_joined_table %>%
   write_csv(here("data", "processed", "beetle_litter_joined_table"))

```
