---
title: "Data Anaysis of Beetle Richness and Leaf Litter"
author: "Zachary krausman, Eryn Lackey, Riley Wares, and Afia Lockett"
date: "2026-03-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load in Packages
```{r load packages}

#install.packages("pacman")
library(pacman)

pacman::p_load(tidyverse, here, cowplot)

```

Import datasets
```{r import datasets}

litter_averages_total <- read.csv(here("data", "processed", "litter_averages_total.csv"))
beetle_abundance_taxonID <- read.csv(here("data", "processed", "beetle_abundance_taxonID.csv"))
beetle_richness_taxonID <- read.csv(here("data", "processed", "beetle_richness_taxonID.csv"))

```

Looking at count of observations per month. 
```{r check monthliness}

ggplot(data=litter_averages_total, aes(x=monthCollected))+
  geom_dotplot(dotsize = 0.025)

ggplot(data=beetle_richness_taxonID, aes(x=month(collectDate)))+
  geom_dotplot(dotsize = 0.025)

```

A note to future explorers: We do not have data for the winter (or between march and october). We are choosing to focus on the rough growing season (april to october)

For discussion: it would be interesting to see what is visible in the winter, but requires more data.

Create average richness with the beetle data. 
```{r create average richness}
beetle_richness_monthly <- beetle_richness_taxonID %>% 
  mutate(monthCollected = month(collectDate), 
         yearCollected = year(collectDate)) %>%
  group_by(plotID, yearCollected, monthCollected) %>%
  summarize(avg_richness_monthly = mean(richness))

```

Preparing beetle and litter data for joining including creating means. 
```{r prelim join tables}
  beetle_richness_by_site_month <- beetle_richness_monthly %>%
  mutate(siteID = substring(plotID,0,4)) %>%
  group_by(siteID, yearCollected, monthCollected) %>%
  summarize(avg_richness_plotmonth = mean(avg_richness_monthly))

  litter_mass_by_site_month <- litter_averages_total %>%
  mutate(siteID = substring(plotID,0,4)) %>%
  group_by(siteID, yearCollected, monthCollected) %>%
  summarize(avg_mass_plotmonth = mean(avgmass), avg_C_plotmonth = mean(avgCpercent), avg_N_plotmonth = mean(avgNpercent))
  
```

Joining litter and beetle data!
```{r}

  beetle_litter_joined_table <- beetle_richness_by_site_month %>%
    full_join(litter_mass_by_site_month) %>%
    drop_na("avg_richness_plotmonth","avg_mass_plotmonth")

```

Dropping C and N because of no overlap with beetle data
```{r drop CN}
BL_beetle_litter_joined_table_clean <- beetle_litter_joined_table %>%
  select(siteID, yearCollected, monthCollected, avg_richness_plotmonth, avg_mass_plotmonth)

```

Scatterplot of beetle richness and average leaf litter mass. 
```{r make a graph}

ggplot(BL_beetle_litter_joined_table_clean, aes(x=avg_mass_plotmonth, y=avg_richness_plotmonth)) +
  geom_point(dotsize=0.5)

```
Exploratory graphing data based on site. 
```{r graph by site}
beetle_litter_model2 <- lm(avg_richness_plotmonth ~ (avg_mass_plotmonth), data=beetle_litter) 
beetle_litter_model <- aov(avg_richness_plotmonth ~ (avg_mass_plotmonth) * siteID, data=beetle_litter) 
summary(beetle_litter_model)
summary(beetle_litter_model2)
plot(beetle_litter_model, which=1:3)

ggplot(beetle_litter_model, aes(x=avg_mass_plotmonth, y=avg_richness_plotmonth, color=siteID)) +
  geom_point(dotsize=0.5) +
  geom_smooth(method="lm")

em <- emmeans(beetle_litter_model, ~siteID)
tukey_beetle_litter <- contrast(em, method="pairwise", adjust="Tukey")
summary(tukey_beetle_litter, infer=TRUE, level=0.95)
```

Export joined beetle richness and average leaf litter data. 
```{r send it out}
BL_beetle_litter_joined_table_clean %>%
   write_csv(here("data", "processed", "beetle_litter_joined_table"))
```
